---
title: "GnRHcell"
output:
  github_document:
    toc: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

<div align="center">

<img src="man/figures/GnRHcell-logo.svg" width="500" alt="GnRHcell logo">

### High-confidence detection, developmental staging, validation, and marker discovery of GnRH neurons from single-cell RNA-seq data

[![R-CMD-check](https://github.com/ymbouamboua/GnRHcell/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/ymbouamboua/GnRHcell/actions/workflows/R-CMD-check.yaml)
[![Reproducibility](https://github.com/ymbouamboua/GnRHcell/actions/workflows/reproducibility.yaml/badge.svg)](https://github.com/ymbouamboua/GnRHcell/actions/workflows/reproducibility.yaml)
[![pkgdown](https://img.shields.io/badge/docs-pkgdown-blue.svg)](https://ymbouamboua.github.io/GnRHcell/)
[![License: MIT](https://img.shields.io/badge/license-MIT-blue.svg)](LICENSE)

</div>

`GnRHcell` is an R package for identifying, characterizing, and validating rare gonadotropin-releasing hormone (GnRH) neurons in single-cell transcriptomic datasets.

It provides:

- high-confidence GnRH-cell detection;
- multi-signal assessment of GnRH identity;
- developmental-stage assignment and stage refinement;
- diagnostic and biological validation;
- GnRH-specific marker discovery and `GNRH1` co-expression analysis;
- multi-dataset processing and comparison;
- conserved cross-dataset marker-program analysis;
- publication-ready visualizations and reports.

## Workflow

```text
Single-cell RNA-seq object
          │
          ▼
 Input validation
          │
          ├── assay and expression-layer checks
          ├── metadata validation
          ├── gene-symbol harmonization
          ├── dimensional-reduction checks
          └── required-feature assessment
          │
          ▼
 GnRH-neuron detection
          │
          ├── GNRH1 expression assessment
          ├── GnRH identity-program scoring
          ├── migration / guidance evidence
          ├── neuroendocrine evidence
          ├── competing-program exclusion
          ├── nearest-neighbor support
          ├── adaptive thresholding
          └── confidence classification
          │
          ▼
 Developmental staging
          │
          ├── neurogenesis
          ├── identity acquisition
          ├── migration
          ├── maturation
          ├── secretion
          └── stage refinement using biological evidence
          │
          ▼
 GnRH-cell characterization
          │
          ├── identity-program activity
          ├── migration-core evidence
          ├── neuroendocrine maturation
          ├── co-expression analysis
          └── GnRH-specific marker discovery
          │
          ▼
 Validation and diagnostics
          │
          ├── detection-class consistency
          ├── score distributions
          ├── developmental-stage composition
          ├── raw-to-final stage transitions
          ├── migration-core validation
          ├── biological marker expression
          └── dataset-level summaries
          │
          ▼
 Multi-dataset analysis
          │
          ├── sequential dataset processing
          ├── runtime comparison
          ├── detected-cell comparison
          ├── marker overlap
          ├── conserved marker programs
          └── cross-dataset validation
          │
          ▼
 Visualization and reporting
          │
          ├── embedding plots
          ├── feature-expression plots
          ├── stage distributions
          ├── co-expression networks
          ├── marker-overlap plots
          └── reproducible reports and tables
```

The package integrates naturally with [Seurat](https://satijalab.org/seurat/) workflows.

Full documentation is available at https://ymbouamboua.github.io/GnRHcell/

## Installation

Install the development version from GitHub:

```{r installation, eval=FALSE}
# install.packages("pak")
pak::pak("ymbouamboua/GnRHcell")
```

Alternatively:

```{r installation-devtools, eval=FALSE}
# install.packages("devtools")
devtools::install_github("ymbouamboua/GnRHcell")
```

## Quick start

```{r quick-start, eval=FALSE}
library(GnRHcell)
library(Seurat)

# `obj` is a normalized Seurat object with a dimensional reduction.
obj <- run_gnrh(obj)
```

The processed Seurat object contains GnRH detection, confidence, evidence scores, and developmental-stage assignments.

```{r inspect-results, eval=FALSE}
table(obj$gnrh_status)
table(obj$gnrh_class)
table(obj$gnrh_confident)
table(obj$gnrh_stage)
```

Additional metadata can include identity, migration, neuroendocrine, exclusion, nearest-neighbor, and stage-refinement metrics.

```{r inspect-scores, eval=FALSE}
obj[[]][
  ,
  c(
    "gnrh_status",
    "gnrh_class",
    "gnrh_support_score",
    "gnrh_identity_score",
    "gnrh_exclusion_score",
    "gnrh_stage"
  )
] |>
  head()
```

## Visualization

### Embedding

```{r embedding, eval=FALSE}
plot_gnrh_embedding(
  obj,
  group.by = c(
    "gnrh_status",
    "gnrh_confident",
    "gnrh_stage"
  ),
  reduction = "umap"
)
```

### Feature expression

```{r features, eval=FALSE}
plot_gnrh_feature(
  obj,
  feature_type = "all",
  reduction = "umap"
)
```

### Distribution across samples

```{r distribution, eval=FALSE}
plot_gnrh_distribution(
  obj,
  group.by = "gnrh_stage",
  split.by = "orig.ident",
  proportion = TRUE,
  label = FALSE,
  cols = gnrh_colors("stage")
)
```

### Diagnostic report

```{r report, eval=FALSE}
gnrh_report(obj)
```

The report summarizes detection signals, threshold performance, GnRH-module activity, classification confidence, and developmental-stage composition.

## GnRH-specific gene discovery

Identify genes enriched in confident GnRH cells relative to a biologically relevant control population:

```{r markers, eval=FALSE}
genes <- find_gnrh_genes(
  object = obj,
  annotation_col = "ann1",
  control_label = "Neuronal",
  donor_col = "status"
)

head(genes$candidates)
```

The returned results combine differential expression, `GNRH1` co-expression, specificity, and donor recurrence.

Consult `?find_gnrh_genes` for available thresholds and returned tables.

### Co-expression and network visualization

```{r marker-plots, eval=FALSE}
coexpressed <- subset(
  genes$markers,
  coexpr_flag %in% TRUE
)

plot_gnrh_coexpr(
  coexpressed,
  coexp_cutoff = 0.3
)

plot_network(
  genes$markers,
  top_n = 50,
  threshold = 0.1
)
```

## Multi-dataset workflow

`GnRHcell` can process collections of independent Seurat datasets sequentially, allowing large studies to be analyzed without loading all objects into memory simultaneously.

### 1. Define the dataset configuration

Each row describes one dataset.

```{r collection-config, eval=FALSE}
datasets <- data.frame(
  id = c(
    "human_hpsc",
    "human_nose",
    "mouse_hypomap"
  ),
  label = c(
    "hPSC GnRH (Wang 2022)",
    "Human fetal nose",
    "Mouse HypoMap"
  ),
  species = c(
    "Human",
    "Human",
    "Mouse"
  ),
  file = c(
    "data/human_hpsc.rds",
    "data/human_nose.rds",
    "data/mouse_hypomap.rds"
  ),
  split_by = c(
    "orig.ident",
    "status",
    "orig.ident"
  ),
  reduction = c(
    "umap",
    "umap",
    "umap_scvi"
  )
)
```

### 2. Validate the configuration

```{r prepare-collection, eval=FALSE}
datasets <- prepare_gnrh_datasets(
  datasets,
  check_files = TRUE,
  remove_missing = FALSE
)

datasets
```

`prepare_gnrh_datasets()` standardizes dataset metadata, expands file paths, checks file availability, detects duplicated IDs, and validates species labels.

### 3. Run GnRHcell across all datasets

```{r run-collection, eval=FALSE}
collection <- run_gnrh_collection(
  datasets = datasets,
  output_dir = "gnrh_results",
  run_markers = TRUE,
  run_comparisons = TRUE,
  run_programs = TRUE,
  clean_objects = TRUE,
  save_objects = FALSE
)
```

The workflow:

- loads datasets sequentially;
- runs `run_gnrh()` on each object;
- generates dataset-level diagnostics and figures;
- identifies GnRH-associated markers;
- writes run-information and marker tables;
- compares datasets;
- optionally identifies conserved marker programs.

The returned object has class `"gnrh_collection"`.

```{r inspect-collection, eval=FALSE}
class(collection)
names(collection$results)
collection$datasets
collection$comparisons
```

### Running a single dataset through the collection workflow

```{r run-single-dataset, eval=FALSE}
result <- run_gnrh_dataset(
  object = obj,
  dataset_id = "human_hpsc",
  dataset_label = "hPSC GnRH (Wang 2022)",
  split_by = "orig.ident",
  reduction = "umap",
  output_dir = "gnrh_results",
  run_markers = TRUE
)

result$run_info
head(result$markers)
```

If the requested dimensional reduction is unavailable, `run_gnrh_dataset()` automatically falls back to `"umap"` when possible.

If the requested splitting variable is absent, commonly used alternatives such as `"orig.ident"`, `"sample"`, and `"library_id"` are considered.

## Cross-dataset comparison

Cross-dataset comparisons can be run automatically by `run_gnrh_collection()` or independently with `compare_gnrh_datasets()`.

```{r compare-datasets, eval=FALSE}
comparison <- compare_gnrh_datasets(
  datasets = datasets,
  output_dir = "gnrh_results",
  run_programs = TRUE
)

comparison$runtime_plot
comparison$detected_plot
comparison$overlap
comparison$programs
```

The comparison workflow can generate:

- runtime comparisons;
- numbers of detected GnRH cells;
- marker-overlap analyses;
- conserved cross-dataset marker programs.

## Cross-dataset validation

`validate_gnrh_collection()` provides an additional validation layer for evaluating GnRH detection and developmental staging across datasets.

Because validation uses the processed Seurat objects, retain them in the collection by setting `clean_objects = FALSE`.

```{r validation-run, eval=FALSE}
collection <- run_gnrh_collection(
  datasets = datasets,
  output_dir = "gnrh_results",
  run_markers = TRUE,
  run_comparisons = TRUE,
  run_programs = TRUE,
  clean_objects = FALSE
)

validation <- validate_gnrh_collection(
  collection
)
```

The returned object has class `"gnrh_validation"` and contains complementary validation tables.

### Detection rates

```{r validation-detection, eval=FALSE}
validation$detection
```

This table reports total cell number, direct detections, supported detections, dropout-rescued detections, total GnRH-positive cells, detection percentages, and the contribution of each detection class.

### Status and class consistency

```{r validation-status-class, eval=FALSE}
validation$status_class
```

This summary evaluates correspondence between `gnrh_status` and `gnrh_class`.

### Evidence-score validation

```{r validation-scores, eval=FALSE}
validation$scores
```

Available GnRH evidence variables are summarized by detection class using medians and interquartile ranges.

### Developmental-stage composition

```{r validation-stages, eval=FALSE}
validation$stages
validation$stage_class
```

These tables summarize developmental stages across all positive GnRH cells and separately within direct, supported, and dropout-rescued detection classes.

### Stage-refinement validation

```{r validation-refinement, eval=FALSE}
validation$stage_refinement
validation$stage_reassignment
```

When stage-refinement metadata are available, these outputs report raw-to-final stage transitions, reassignment frequencies, and reasons for stage refinement.

### Migration-core validation

```{r validation-migration-core, eval=FALSE}
validation$migration_core
```

Migration-core evidence is summarized by raw developmental stage using median and interquartile-range marker hits together with the percentage of cells having at least 1, 2, or 3 migration-core hits.

### Independent biological marker validation

```{r validation-biological-markers, eval=FALSE}
validation$biological_markers
```

By default, expression is evaluated for a curated panel spanning GnRH identity, specification, migration, guidance, and neuroendocrine maturation:

```text
GNRH1, GNRHR,
FEZF1, ISL1, OTX2, SIX3, SIX6,
CHGA, CHGB, SCG2, PCSK1, PCSK2, CPE, VGF, SYP,
ANOS1, PROKR2, PROK2, NRP1, NRP2, ROBO1, ROBO2, DCX
```

For each detection class, `validate_gnrh_collection()` reports average expression and the percentage of cells expressing each available marker.

A custom panel can also be supplied:

```{r validation-custom-markers, eval=FALSE}
validation <- validate_gnrh_collection(
  collection,
  validation_markers = c(
    "GNRH1",
    "FEZF1",
    "ISL1",
    "ANOS1",
    "PROKR2",
    "DCX",
    "CHGA",
    "PCSK2"
  )
)
```

Validation tables are written by default to:

```text
gnrh_results/
└── validation/
    ├── input_summary.csv
    ├── detection.csv
    ├── status_class.csv
    ├── scores.csv
    ├── stages.csv
    ├── stage_class.csv
    ├── stage_refinement.csv
    ├── stage_reassignment.csv
    ├── migration_core.csv
    └── biological_markers.csv
```

## Cross-dataset marker programs

Marker-program analysis can also be performed manually from previously generated marker files.

```{r marker-programs, eval=FALSE}
files <- c(
  "HuDeCa Nose" = "gnrh_human_nose_markers.tsv",
  "HPSC Wang 2022" = "gnrh_human_hpsc_markers.tsv",
  "Human HypoMap" = "gnrh_human_hypomap_markers.tsv"
)

marker_dir <- file.path(outdir, "markers")

gene_sets <- build_gene_sets(
  files = files,
  dir = marker_dir
)

overlap <- gene_upset(
  gene_sets = gene_sets,
  outdir = file.path(
    outdir,
    "comparisons",
    "marker_overlap"
  )
)

programs <- gnrh_marker_programs(
  files = files,
  results = overlap,
  outdir = outdir
)

head(programs$candidate_table)
head(programs$high_confidence)
```

### Marker-program visualization

```{r marker-program-plots, eval=FALSE}
plot_gnrh_marker_programs(
  programs,
  table = "summary",
  type = "bar"
)

plot_gnrh_marker_programs(
  programs,
  table = "summary",
  type = "tile"
)
```

## Output structure

A typical multi-dataset run produces:

```text
gnrh_results/
├── figures/
│   ├── human_hpsc/
│   │   ├── human_hpsc_gnrh_report.pdf
│   │   ├── human_hpsc_gnrh_embedding.pdf
│   │   ├── human_hpsc_gnrh_features.pdf
│   │   ├── human_hpsc_status_distribution.pdf
│   │   ├── human_hpsc_stage_distribution.pdf
│   │   ├── human_hpsc_coexpression.pdf
│   │   └── human_hpsc_network.pdf
│   └── ...
├── tables/
│   ├── human_hpsc_gnrh_run_info.tsv
│   └── ...
├── markers/
│   ├── gnrh_human_hpsc_markers.tsv
│   └── ...
├── comparisons/
│   ├── marker_overlap/
│   └── marker_programs/
├── validation/
│   ├── input_summary.csv
│   ├── detection.csv
│   ├── status_class.csv
│   ├── scores.csv
│   ├── stages.csv
│   ├── stage_class.csv
│   ├── stage_refinement.csv
│   ├── stage_reassignment.csv
│   ├── migration_core.csv
│   └── biological_markers.csv
└── objects/
    └── human_hpsc_gnrh.rds
```

The `objects/` directory is populated only when `save_objects = TRUE`.

## Main functions

### Core detection and staging

- `run_gnrh()`
- `detect_gnrh()`
- `stage_gnrh()`
- `gnrh_diagnostics()`

### Dataset collections

- `prepare_gnrh_datasets()`
- `run_gnrh_dataset()`
- `run_gnrh_collection()`
- `compare_gnrh_datasets()`
- `validate_gnrh_collection()`

### Marker analysis

- `find_gnrh_genes()`
- `gnrh_markers()`
- `build_gene_sets()`
- `gene_upset()`
- `gnrh_marker_programs()`

### Visualization and reporting

- `plot_gnrh_embedding()`
- `plot_gnrh_feature()`
- `plot_gnrh_distribution()`
- `plot_gnrh_coexpr()`
- `plot_network()`
- `plot_gnrh_marker_programs()`
- `plot_gnrh_runtime_curve()`
- `plot_gnrh_detected()`
- `gnrh_report()`

### Utilities

- `gnrh_colors()`
- `cellpal()`
- `plot_theme()`

See the [function reference](https://ymbouamboua.github.io/GnRHcell/reference/) for the complete API.

## Recommended multi-dataset analysis

For a complete cross-dataset analysis with validation:

```{r complete-workflow, eval=FALSE}
library(GnRHcell)

datasets <- prepare_gnrh_datasets(
  datasets,
  check_files = TRUE
)

collection <- run_gnrh_collection(
  datasets = datasets,
  output_dir = "gnrh_results",
  run_markers = TRUE,
  run_comparisons = TRUE,
  run_programs = TRUE,
  clean_objects = FALSE,
  save_objects = FALSE
)

validation <- validate_gnrh_collection(
  collection
)

validation$detection
validation$scores
validation$stages
validation$stage_refinement
validation$migration_core
validation$biological_markers
```

If memory usage is a concern, processed objects can instead be saved during the main analysis:

```{r memory-workflow, eval=FALSE}
collection <- run_gnrh_collection(
  datasets = datasets,
  output_dir = "gnrh_results",
  clean_objects = TRUE,
  save_objects = TRUE
)
```

This mode minimizes memory usage during large multi-dataset studies.

## Development

```{r development, eval=FALSE}
devtools::document()
devtools::test()
devtools::check()
pkgdown::build_site()
```

Rebuild `README.md` after editing this source file:

```{r rebuild-readme, eval=FALSE}
devtools::build_readme()
```

## Citation

If you use `GnRHcell`, please cite:

> Mbouamboua Y. *GnRHcell: high-confidence GnRH neuron detection, developmental staging, and cross-dataset validation from single-cell RNA-seq data.*

## License

`GnRHcell` is available under the [MIT License](LICENSE).

## Development status

`GnRHcell` is under active development. Interfaces and biological models may evolve as detection, staging, and cross-dataset validation methods are refined.
