---
title: "GnRHcell"
output:
  github_document:
    toc: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

<div align="center">

<img src="man/figures/GnRHcell-logo.svg" width="500" alt="GnRHcell logo">


### High-confidence detection, developmental staging, and marker discovery of GnRH neurons from single-cell RNA-seq data

[![R-CMD-check](https://github.com/ymbouamboua/GnRHcell/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/ymbouamboua/GnRHcell/actions/workflows/R-CMD-check.yaml)
[![Reproducibility](https://github.com/ymbouamboua/GnRHcell/actions/workflows/reproducibility.yaml/badge.svg)](https://github.com/ymbouamboua/GnRHcell/actions/workflows/reproducibility.yaml)
[![pkgdown](https://img.shields.io/badge/docs-pkgdown-blue.svg)](https://ymbouamboua.github.io/GnRHcell/)
[![License: MIT](https://img.shields.io/badge/license-MIT-blue.svg)](LICENSE)

</div>

`GnRHcell` is an R package for identifying rare gonadotropin-releasing hormone (GnRH) neurons in single-cell transcriptomic datasets.

It provides:

- high-confidence GnRH-cell detection;
- developmental-stage assignment;
- diagnostic performance evaluation;
- marker discovery and `GNRH1` co-expression analysis;
- cross-dataset marker-program analysis;
- publication-ready visualizations.

The package integrates naturally with [Seurat](https://satijalab.org/seurat/) workflows. Full documentation is available at [ymbouamboua.github.io/GnRHcell](https://ymbouamboua.github.io/GnRHcell/).


## Installation

Install the development version from GitHub:

```{r installation, eval=FALSE}
# install.packages("pak")
pak::pak("ymbouamboua/GnRHcell")
```

Alternatively:

```{r installation-devtools, eval=FALSE}
# install.packages("devtools")
devtools::install_github("ymbouamboua/GnRHcell")
```

## Quick start

```{r quick-start, eval=FALSE}
library(GnRHcell)
library(Seurat)

# `obj` is a normalized Seurat object with a dimensional reduction.
obj <- run_gnrh(obj)
```

The resulting metadata include GnRH detection, confidence, and developmental-stage assignments.

```{r inspect-results, eval=FALSE}
table(obj$gnrh_status)
table(obj$gnrh_confident)
table(obj$gnrh_stage)
```

## Visualization

### Embedding

```{r embedding, eval=FALSE}
plot_gnrh_embedding(
  obj,
  group.by = c("gnrh_status", "gnrh_confident", "gnrh_stage"),
  reduction = "umap"
)
```

### Feature expression

```{r features, eval=FALSE}
plot_gnrh_feature(obj, feature_type = "all", reduction = "umap")
```

### Distribution across samples

```{r distribution, eval=FALSE}
plot_gnrh_distribution(
  obj,
  group.by = "gnrh_stage",
  split.by = "orig.ident",
  proportion = TRUE,
  label = FALSE,
  cols = gnrh_colors("stage")
)
```

### Diagnostic report

```{r report, eval=FALSE}
gnrh_report(obj)
```

The report summarizes signal distributions, threshold performance, module activity, stage composition, and classification diagnostics.

## GnRH-specific gene discovery

Identify genes enriched in confident GnRH cells relative to a biologically relevant control population:

```{r markers, eval=FALSE}
genes <- find_gnrh_genes(
  object = obj,
  annotation_col = "ann1",
  control_label = "Neuronal",
  donor_col = "status"
)

head(genes$candidates)
```

The returned results combine differential expression, GNRH1 co-expression, specificity, and donor recurrence. Consult `?find_gnrh_genes` for the available thresholds and returned tables.

### Co-expression and network visualization

```{r marker-plots, eval=FALSE}
coexpressed <- subset(genes$markers, coexpr_flag %in% TRUE)

plot_gnrh_coexpr(coexpressed, coexp_cutoff = 0.3)
plot_network(genes$markers, top_n = 50, threshold = 0.1)
```

## Cross-dataset marker programs

```{r marker-programs, eval=FALSE}
files <- c(
  "HuDeCa Nose" = "gnrh_human_nose_markers.tsv",
  "HPSC Wang 2022" = "gnrh_human_hpsc_markers.tsv",
  "Human HypoMap" = "gnrh_human_hypomap_markers.tsv"
)

marker_dir <- file.path(outdir, "markers")
gene_sets <- build_gene_sets(files = files, dir = marker_dir)

overlap <- gene_upset(
  gene_sets = gene_sets,
  outdir = file.path(outdir, "comparisons", "marker_overlap")
)

programs <- gnrh_marker_programs(
  files = files,
  results = overlap,
  outdir = outdir
)

head(programs$candidate_table)
head(programs$high_confidence)
```

```{r marker-program-plots, eval=FALSE}
plot_gnrh_marker_programs(programs, table = "summary", type = "bar")
plot_gnrh_marker_programs(programs, table = "summary", type = "tile")
```

## Main functions

- Pipeline: `run_gnrh()`, `detect_gnrh()`, `stage_gnrh()`, `gnrh_diagnostics()`
- Marker analysis: `find_gnrh_genes()`, `gnrh_markers()`
- Cross-dataset analysis: `build_gene_sets()`, `gene_upset()`, `gnrh_marker_programs()`
- Visualization: `plot_gnrh_embedding()`, `plot_gnrh_feature()`, `plot_gnrh_distribution()`, `gnrh_report()`
- Utilities: `gnrh_colors()`, `cellpal()`, `plot_theme()`

See the [function reference](https://ymbouamboua.github.io/GnRHcell/reference/) for the complete API.

## Development

```{r development, eval=FALSE}
devtools::document()
devtools::test()
devtools::check()
pkgdown::build_site()
```

Rebuild `README.md` after editing this source file:

```{r rebuild-readme, eval=FALSE}
devtools::build_readme()
```

## Citation

If you use `GnRHcell`, please cite:

> Mbouamboua Y. *GnRHcell: high-confidence GnRH neuron detection and developmental staging from single-cell RNA-seq data.*

## License

`GnRHcell` is available under the [MIT License](LICENSE).

## Development status

`GnRHcell` is under active development. Interfaces may evolve as the methods are refined.
