---
title: "GnRHcell: GnRH neuron identification and developmental staging"
author:
  - name: Yvon Mbouamboua
  - name: Paolo Giacobini
output:
  rmarkdown::html_vignette:
    toc: true
    number_sections: false
vignette: >
  %\VignetteIndexEntry{GnRHcell: GnRH neuron identification and developmental staging}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE,
  fig.align = "center",
  fig.width = 8,
  fig.height = 6
)
```

# Overview

`GnRHcell` is an R package for high-confidence detection, developmental staging, diagnostics, and marker discovery of gonadotropin-releasing hormone (GnRH) neurons from single-cell RNA-seq data.

GnRH neurons are rare hypothalamic neuroendocrine cells essential for reproductive function. In scRNA-seq datasets, their identification is difficult because of:

- low or sparse `GNRH1` expression
- transcriptional dropout
- ambient RNA contamination
- developmental heterogeneity
- rare-cell abundance
- incomplete neuroendocrine maturation signatures

`GnRHcell` is designed for Seurat-based workflows and integrates direct `GNRH1` expression, GnRH lineage markers, migration-associated programs, neuroendocrine maturation markers, neighborhood support, adaptive thresholding, diagnostics, and developmental stage inference.

# Installation

```{r eval=FALSE}
# install.packages("remotes")
remotes::install_github("ymbouamboua/GnRHcell")
```

```{r eval=FALSE}
library(GnRHcell)
library(Seurat)
```

# Input requirements

`GnRHcell` expects a Seurat object containing:

- raw UMI counts
- gene names
- cell barcodes
- a valid RNA assay
- normalized data for plotting and downstream visualization

A typical object should contain a `counts` layer/slot and preferably a `data` layer/slot.

```{r eval=FALSE}
DefaultAssay(obj)
dim(obj)
```

# Example dataset

As an example, the package can be applied to the scRNA-seq dataset from Wang et al. (2022), which profiled human pluripotent stem cell-derived GnRH neuron differentiation cultures. The count matrices are available through GEO accession `GSE212901`.

```{r eval=FALSE}
samples <- c("D22", "D24", "D26")
dir.data <- "path/to/Wang_GSE212901_RAW"

seurat.list <- lapply(samples, function(x) {
  counts <- Read10X(file.path(dir.data, x))
  CreateSeuratObject(counts = counts, project = x)
})

names(seurat.list) <- samples

obj <- merge(seurat.list[[1]], y = seurat.list[-1])
obj[["percent.mt"]] <- PercentageFeatureSet(obj, pattern = "^MT-")

obj <- subset(
  obj,
  subset = nFeature_RNA > 1000 &
    nCount_RNA > 1500 &
    percent.mt < 20
)

obj <- NormalizeData(obj)
obj <- FindVariableFeatures(obj)
obj <- ScaleData(obj)
obj <- RunPCA(obj)
obj <- RunUMAP(obj, dims = 1:30)
```

# Run the complete GnRHcell pipeline

The main entry point is `run_gnrh()`.

```{r eval=FALSE}

data(hpsc_gnrh)

hpsc_gnrh <- run_gnrh(hpsc_gnrh)

```

The pipeline runs:

1. GnRH neuron detection with `detect_gnrh()`
2. developmental stage assignment with `stage_gnrh()`
3. diagnostic analysis with `gnrh_diagnostics()`
4. runtime and summary reporting

Example console output:

```text
[GNRH] ==== STARTING GnRHcell PIPELINE ====
[STEP] [1/3] Detecting GnRH cells
[DONE] Detection complete. Duration: 7.6s
[STEP] [2/3] Assigning developmental stages
[DONE] Staging complete. Duration: 2.3s
[INFO] [3/3] Running diagnostics
[DONE] Diagnostics complete. Duration: 1.1s
[INFO] PIPELINE SUMMARY
[INFO] Status:
[INFO]   neg: 28452
[INFO]   pos: 317
[INFO] Stage:
[INFO]   identity: 62
[INFO]   migrating: 121
[INFO]   mature: 134
[INFO]   secreting: 0
[INFO]   non-gnrh: 28452
[DONE] ==== GnRHcell PIPELINE COMPLETE ==== Duration: 11.2s
```

# Detection model

`detect_gnrh()` uses a composite evidence framework combining:

- direct `GNRH1` expression
- core GnRH lineage markers
- migration and axon-guidance markers
- neuroendocrine maturation markers
- k-nearest-neighbor support
- adaptive score thresholding

An ambient RNA ratio is stored for diagnostics only; it does not correct expression or affect classification.

The default detection logic is designed to reduce false positives caused by isolated `GNRH1` transcripts while retaining rare GnRH-like cells with broader biological support.

# Marker modules

## Core GnRH identity module

The core module contains developmental and lineage-associated genes such as:

- `GNRH1`
- `FEZF1`
- `ISL1`
- `OTX2`
- `SIX3`
- `SIX6`
- `DLX1`
- `DLX2`
- `DLX5`
- `DLX6`

## Migration module

GnRH neurons originate near the olfactory placode and migrate toward the hypothalamus. Migration-related markers include:

- `ANOS1`
- `PROKR2`
- `PROK2`
- `NRP1`
- `NRP2`
- `SEMA3A`
- `SEMA3C`
- `SEMA3F`
- `ROBO1`
- `ROBO2`
- `L1CAM`
- `DCX`

## Neuroendocrine module

Neuroendocrine maturation markers include:

- `KISS1R`
- `TAC3`
- `TACR3`
- `GNRHR`
- `PCSK1`
- `PCSK2`
- `SCG2`
- `CHGA`
- `CHGB`
- `CPE`
- `VGF`
- `SYP`
- `RAB3A`


# Composite score

For each cell, `GnRHcell` computes a composite score integrating
expression and marker support:

$$
S_i = 3G_i + I_i + M_i + N_i + H_i + 0.75K_i
$$

where $G_i$ is log-normalized `GNRH1` expression; $I_i$, $M_i$,
$N_i$, and $H_i$ are the weighted identity, migration, neuroendocrine,
and hormone module scores; and $K_i$ is neighborhood identity support.
The raw score is stored as `gnrh_score_raw` and standardized across cells as
`gnrh_score` for reporting.

Classification uses direct GNRH1 counts and independent identity evidence.
The separate support score excludes GNRH1 and ambient RNA:

$$
U_i = 4I_i + 0.35M_i + 0.75N_i + 0.10H_i + 0.35K_i.
$$

# Adaptive thresholding

Expression and score thresholds are estimated adaptively from each
dataset. The expression threshold is estimated from non-zero normalized
`GNRH1` expression using:

$$
T = \operatorname{median}(x) +
\lambda\,\operatorname{MAD}(x)
$$

where $x$ represents non-zero normalized `GNRH1` expression and
$\lambda$ is controlled by `mad_factor`. A quantile-based fallback is
used when robust estimation is not possible. This expression threshold is
diagnostic only. Supported low-UMI calls instead use a quantile of the raw
support scores among trusted direct-reference cells, controlled by
`supported_q`. They are disabled when fewer than `min_reference_cells`
trusted reference cells are available.


# Developmental staging

After detection, `stage_gnrh()` separates GnRH lineage identity from the
developmental interpretation. The preferred output is `gnrh_stage`.

Current stages are:

- `early`
- `migrating`
- `post-migratory`
- `mature`
- `transitional`
- `undetermined`
- `non-gnrh`

```{r eval=FALSE}

table(hpsc_gnrh$gnrh_stage)
```

Secretory support is reported independently in `gnrh_secretory`.

# Diagnostic report

`gnrh_report()` generates a multi-panel diagnostic dashboard.

```{r eval=FALSE}

p <- gnrh_report(hpsc_gnrh)
p

```

The report summarizes:

- `GNRH1` expression versus GnRH score
- score distributions
- threshold performance curve
- ROC analysis when truth labels are available
- module hit enrichment
- GnRH class/status composition

# Embedding visualization

Plot GnRH status and stage on a Seurat embedding:

```{r eval=FALSE}

p <- gnrh_cellmap(hpsc_gnrh, group_by = "gnrh_status") |
  gnrh_cellmap(hpsc_gnrh, group_by = "gnrh_stage")

p

```

# Feature visualization

Plot GnRH diagnostic features:

```{r eval=FALSE}

p <- gnrh_cellfeat(
  hpsc_gnrh,
  preset = "all"
)

p

```

Common GnRH diagnostic features include:

- `gnrh_raw`
- `gnrh_expr`
- `gnrh_score`
- `gnrh_knn`
- `gnrh_core_hits`
- `gnrh_mig_hits`
- `gnrh_neuro_hits`

# Distribution plots

GnRH status by sample:

```{r eval=FALSE}

gnrh_celldistribution(
  hpsc_gnrh,
  group.by = "gnrh_status",
  split.by = "orig.ident",
  proportion = TRUE,
  label = FALSE,
  cols = gnrh_palette("status")
)

```

GnRH stage by sample:

```{r eval=FALSE}

gnrh_celldistribution(
  hpsc_gnrh,
  group.by = "gnrh_stage",
  split.by = "orig.ident",
  proportion = TRUE,
  label = FALSE,
  cols = gnrh_palette("stage")
)

```

# Marker discovery

Identify GnRH-associated markers:

```{r eval=FALSE}

markers <- gnrh_markers(hpsc_gnrh)
head(markers)

```

The marker table may include:

- differential expression statistics
- detection frequency
- specificity metrics
- coexpression with `GNRH1`
- composite marker scores
- coexpression flags


# GnRH-specific gene discovery

`find_gnrh_genes()` identifies genes enriched in confident GnRH cells
relative to a biologically relevant control population. It combines
differential expression, `GNRH1` co-expression, cellular specificity,
and recurrence across donors.

```{r marker-discovery, eval=FALSE}
genes <- find_gnrh_genes(
  object = hpsc_gnrh,
  annotation_col = "ann1",
  control_label = "Neuronal",
  donor_col = "orig.ident"
)

head(genes$candidates)
```

The returned object contains:

- differential-expression results;
- `GNRH1` co-expression frequencies;
- specificity scores;
- donor-level detection frequencies;
- prioritized candidate genes.

The lower-level `gnrh_markers()` function remains available when only
the original marker table is required.

```{r legacy-markers, eval=FALSE}
markers <- gnrh_markers(hpsc_gnrh)
head(markers)
```


# Coexpression visualization

Filter coexpressed markers:

```{r eval=FALSE}

df <- subset(markers, coexpr_flag == TRUE)

```

Plot top coexpressed genes:

```{r eval=FALSE}

gnrh_coexpr(
  df,
  coexp_cutoff = 0.3
)

```

# Gene network

Build a marker similarity network:

```{r eval=FALSE}

gnrh_network(
  markers,
  top_n = 50,
  threshold = 0.1
)

```


# Stored metadata

After running `run_gnrh()`, the Seurat object contains additional metadata columns such as:

```{r eval=FALSE}

grep("^gnrh", colnames(hpsc_gnrh@meta.data), value = TRUE)

```

Common fields include:

- `gnrh_status`
- `gnrh_stage`
- `gnrh_score`
- `gnrh_expr`
- `gnrh_raw`
- `gnrh_core_hits`
- `gnrh_mig_hits`
- `gnrh_neuro_hits`
- `gnrh_knn`
- `gnrh_truth`

Runtime and diagnostic information are stored in:

```{r eval=FALSE}
names(hpsc_gnrh@misc$gnrh)
```

# Recommended workflow

```{r eval=FALSE}

obj <- NormalizeData(obj)
obj <- FindVariableFeatures(obj)
obj <- ScaleData(obj)
obj <- RunPCA(obj)
obj <- RunUMAP(obj, dims = 1:30)
obj <- run_gnrh(obj)
gnrh_report(obj)
gnrh_cellmap(obj, group_by = "gnrh_status")
gnrh_cellmap(obj, group_by = "gnrh_stage")
gnrh_cellfeat(obj, preset = "all")
markers <- gnrh_markers(obj)
df <- subset(markers, coexpr_flag == TRUE)
gnrh_coexpr(df, coexp_cutoff = 0.3)
gnrh_network(markers, top_n = 50, threshold = 0.1)

```

# Best practices

For droplet-based scRNA-seq, a typical starting configuration is:

```{r eval=FALSE}

obj <- run_gnrh(
  obj,
  min_umi = 2,
  min_counts = 500,
  mad_factor = 2,
  score_q = 0.9
)

```

For lower-depth datasets, consider relaxing thresholds:

```{r eval=FALSE}

obj <- run_gnrh(
  obj,
  min_umi = 1,
  min_counts = 200,
  score_q = 0.75
)

```

All threshold choices should be validated using the diagnostic report and biological marker context.

# Reproducibility

`GnRHcell` stores:

- detection parameters
- thresholds
- module scores
- diagnostic tables
- stage assignments
- runtime information
- session information

This supports reproducible downstream analysis.

# Session information

```{r eval=FALSE}

sessionInfo()

```
