---
title: "An integrated framework for GnRH neuron identification and developmental trajectory inference"
author: "Yvon Mbouamboua & Paolo Giacobini"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{GnRHcell}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
# Overview

Gonadotropin-releasing hormone (GnRH) neurons are rare hypothalamic neuroendocrine cells essential for reproductive function. In single-cell RNA sequencing (scRNA-seq) datasets, their identification is challenging due to:

* low expression of GNRH1
* transcriptional dropout
* ambient RNA contamination
* developmental heterogeneity
* sparse transcriptomic profiles

The `cellgnrh` R package provides a biologically informed framework for identifying and staging GnRH-expressing neurons from scRNA-seq datasets.

The workflow integrates:

* sparse-aware normalization
* developmental lineage scoring
* migration signature detection
* neuroendocrine maturation assessment
* ambient RNA correction
* adaptive thresholding
* confidence classification
* developmental stage inference

The framework is specifically designed for rare neuronal populations where conventional marker-based filtering approaches often fail.

---

# Installation

```{r eval=FALSE}
#remotes::install_github("ymbouamboua/cellgnrh")
```

# Load required libraries :

```{r}
# library(cellgnrh)
# library(Seurat)
# library(ggplot2)
```


Input Requirements

The workflow expects a Seurat object containing:

* raw UMI counts
* gene names
* cell barcodes

Typical assay:

```{r}
#DefaultAssay(obj)
```

# Example dataset

To illustrate the methodology, we use the scRNA-seq dataset from Yafei Wang et al. (2022), comprising 29,623 cells from human pluripotent stem cell-derived GnRH neuron differentiation cultures. The count matrices are publicly available through Gene Expression Omnibus under accession number GSE212901. 

```{r}

# # Load Wang et al. scRNA-seq data
# samples <- c("D22", "D24", "D26")
# dir.data <- "/Users/yvon.mbouamboua/projects/singlecell/10x-data/data/Wang_GSE212901_RAW"
# 
# seurat.list <- lapply(samples, function(x) {
#   counts <- Read10X(file.path(dir.data, x))
#   CreateSeuratObject(counts = counts, project = x)
# })
# names(seurat.list) <- samples
# 
# # Quality control
# obj <- merge(seurat.list[[1]], y = seurat.list[-1])
# obj[["percent.mt"]] <- PercentageFeatureSet(obj, pattern = "^MT-")
# 
# obj <- subset(
#   obj,
#   subset = nFeature_RNA > 1000 &
#            nCount_RNA > 1500 &
#            percent.mt < 20
# )
# 
# # Standard preprocessing and integration
# obj <- NormalizeData(obj)
# obj <- FindVariableFeatures(obj)
# obj <- ScaleData(obj)
# obj <- RunPCA(obj)
# 
# obj <- IntegrateLayers(
#   object = obj,
#   method = HarmonyIntegration,
#   orig.reduction = "pca",
#   new.reduction = "harmony"
# )
# 
# # JoinLayers
# obj <- JoinLayers(obj)
# 
# # Clustering and visualization
# obj <- FindNeighbors(obj, reduction = "harmony", dims = 1:30)
# obj <- FindClusters(obj, resolution = 0.2)
# obj <- RunUMAP(obj, reduction = "harmony", dims = 1:30, min.dist = 0.3, spread = 0.5)
# DimPlot(obj, group.by = "seurat_clusters")
# 
# # Save integrated data
# saveRDS(obj, file.path(dir.data, "wang.rds"))

```


Inspect dimensions:

```{r}
#dim(obj)
```


# Step 1 — Detect GnRH-expressing cells

Run the main detection pipeline:

```{r}
# obj <- gnrh_detect(
#   obj,
#   assay = "RNA",
#   layer = "counts",
#   min_umi = 2,
#   mad_factor = 2,
#   score_q = 0.85,
#   min_counts = 500,
#   verbose = TRUE
# )
```

## Biological logic of the pipeline

The detection framework integrates multiple biological modules.


### Core developmental identity markers

The core developmental module captures transcription factors and developmental regulators known to participate in hypothalamic specification and GnRH neuronal differentiation.

| Gene | Biological role |  Key references |
|--------------|--------------------|----------------------------------------|
| GNRH1 | Encodes gonadotropin-releasing hormone; defining marker of GnRH neurons | Wray S. (2001); Herbison AE. (2018) |
| FEZF1 | Required for olfactory system development and GnRH neuron migration|Kotan et al. (2014)|
| ISL1 | LIM-homeodomain factor involved in neuronal differentiation | Catela et al. (2016) |
|OTX2|Controls forebrain and hypothalamic development|Diaczok et al. (2011)|
|SIX3/SIX6|Forebrain patterning and hypothalamic specification|Larder et al. (2011)|
|DLX1/2/5/6|GABAergic and olfactory placode developmental programs|Long et al. (2009)|



### Migration program markers

GnRH neurons originate in the olfactory placode and migrate toward the hypothalamus during embryogenesis.

| Gene | Biological role |  Key references |
|--------------|--------------------|----------------------------------------|
|ANOS1|Axonal guidance and neuronal migration; associated with Kallmann syndrome|Hardelin et al. (1992)|
|PROKR2/PROK2|Regulate migratory guidance pathways|Pitteloud et al. (2007)|
|NRP1/NRP2|Semaphorin receptors involved in axon guidance|Cariboni et al. (2007)|
|SEMA3A/SEMA3C/SEMA3F|Guidance cues for olfactory and GnRH migration|Young et al. (2012)|
|ROBO1/ROBO2|Slit-Robo migratory signaling|Andrews et al. (2006)|
|L1CAM|Neuronal adhesion and migration|Schwarting et al. (2004)|
|DCX|Immature migrating neuron marker|Gleeson et al. (1999)|



### Neuroendocrine maturation markers

These genes characterize differentiated neurosecretory neuronal states.

| Gene | Biological role |  Key references |
|--------------|--------------------|----------------------------------------|
|KISS1R|Puberty and reproductive axis regulation|Seminara et al. (2003)|
|TAC3/TACR3|Neurokinin B signaling in reproductive control|Topaloglu et al. (2009|
|GNRHR|GnRH receptor signaling|Millar et al. (2004)|
|PCSK1/PCSK2|Prohormone processing enzymes|Seidah and Chretien (1999)|
|CHGA/CHGB|Neurosecretory granule markers|Helle et al. (2007)|
|SCG2|Secretory granule protein|Bartolomucci et al. (2011)|
|CPE|Peptide processing and vesicle trafficking|Fricker (1988)|
|VGF|Neuroendocrine secretory pathway marker|Levi et al. (2004)|
|SYP/RAB3A|Synaptic vesicle biology|Sudhof (1995)|

## Sparse-aware composite scoring

The GnRH confidence score integrates multiple biological modules:


$S_i = w_1 G_i + w_2 C_i + w_3 M_i + w_4 N_i$

where:

* $S_i$ = composite GnRH score for cell i
* $G_i$ = normalized GNRH1 expression
* $C_i$ = developmental core module score
* $M_i$ = migration module score
* $N_i$ = neuroendocrine maturation score

Default weights:

$S_i = 2.5G_i + 2.0C_i + 1.0M_i + 1.5N_i$

The weighting scheme prioritizes direct GNRH1 detection while retaining biologically informative auxiliary programs.


## Module score computation

For a module containing k genes:

$M_{module,i} = \frac{1}{k} \sum_{j=1}^{k} E_{ij}$

where:

* $E_{ij}$ is normalized expression of gene $j$ in cell $i$
* module scores are computed after library-size normalization and log transformation.

Sparse-aware averaging reduces instability caused by transcriptional dropout.


## Adaptive thresholding

Primary thresholds are estimated using robust statistics:

$T = \mathrm{median}(S) + \lambda \times MAD(S)$

where:

* $MAD$ is the median absolute deviation
* $\lambda$ corresponds to the user-defined mad_factor

Fallback quantile thresholding:

$T_q = Q_q(S)$

where:

* $Q_q$ is the score quantile
* $q$ corresponds to parameter $score_q$

This hybrid strategy stabilizes classification in highly sparse datasets.


## Ambient RNA correction

Ambient contamination is estimated from low-complexity droplets:

$A_g = \frac{1}{n} \sum_{i \in D} E_{ig}$

where:

* $A_g$ is estimated ambient abundance for gene $g$
* $D$ represents low-UMI droplets

Corrected expression:

$E'_{ig} = \max(E_{ig} - A_g, 0)$

This procedure reduces false-positive detection caused by ambient GNRH1 transcripts.


## Developmental stage assignment

Cells are assigned to developmental states using dominant biological programs.

For stage-specific module scores:

$Stage_i = \arg\max_k (M_{ik})$

where:

* $M_{ik}$ is the score of stage program $k$ in cell $i$

Stages represent progressive developmental transitions:

$progenitor \rightarrow migratory \rightarrow transitional \rightarrow mature \rightarrow secretory$

## Adaptive thresholding

Thresholds are determined dynamically using:

* median + MAD
* quantile fallback
* composite confidence filtering

This allows robust detection under sparse expression regimes.


The final confidence score integrates:

* direct GNRH1 expression
* developmental identity
* migration signatures
* neuroendocrine maturation

The weighted model:

```{r eval=FALSE}
# score =
#   2.5 * GNRH1 +
#   2.0 * Core +
#   1.0 * Migration +
#   1.5 * Neuroendocrine
```


## Cell classification

Cells are classified into:

| Class | Description |
|--------------|--------------------|
| gnrh_neg | no detectable GnRH identity | 
| gnrh_low|weak/intermediate evidence|
| gnrh_high | high-confidence GnRH neuron |

## Detection results

Inspect classification counts:

```{r}
#table(obj$gnrh_class)
```

Detected lineage

```{r}
#table(obj$gnrh_detected)
```

## Visualization

UMAP Classification

```{r}
#plot_gnrh_umap(obj)
```

Threshold Diagnostics

```{r}
#plot_gnrh_threshold(obj)
```

Threshold Optimization Curve

```{r}
#plot_gnrh_threshold_curve(obj)
```

ROC Analysis

```{r}
#plot_gnrh_roc(obj)
```

Feature Visualization

```{r}
# plot_gnrh_feature(
#   obj,
#   gene = c("gnrh_expr", "gnrh_score")
# )
```


Violin plot

```{r}
#plot_gnrh_violin(obj)
```


# Step 2 — Developmental staging

After lineage detection, cells can be assigned developmental states.

```{r}
#obj <- gnrh_stage(obj)
```

## GnRH developmental stages

The staging framework models progressive maturation.

| Class | Description |
|--------------|--------------------|
| progenitor | early neural progenitors | 
| migratory|active migratory neurons|
| transitional | lineage transition state |
|mature|differentiated neuroendocrine neurons|
|secretory|hormone-secreting state|
|other|non-GnRH lineage|


## Probabilistic classification

Instead of hard thresholding only, this estimates the probability that each cell belongs to the GnRH lineage.

This is highly valuable for:

* transitional states
* developmental continua
* low-depth datasets
* noisy droplet data

$P(\mathrm{GnRH}_i) = \frac{1}{1 + e^{-k(S_i - \mu)}}$

where:

* $S_i$ is the composite GnRH score
* $\mu$ is the score midpoint
* $k$ controls classification sharpness


```{r}
# obj = gnrh_probability(
#   obj,
#   score_col = "gnrh_score",
#   midpoint = NULL,
#   slope = 2,
#   threshold = 0.5,
#   verbose = TRUE
# )
```


## Stage visualization

UMAP Staging

```{r}
# plot_gnrh_umap(
#   obj,
#   group.by = "gnrh_stage"
# )
```

Stage Score Distributions

```{r}
#plot_gnrh_stage_scores(obj)
```


## Marker gene identification

```{r}

# markers <- gnrh_markers(
#   obj,
#   group.by = "gnrh_predicted",
#   ident.1 = gnrh_pos,
#   ident.2 = gnrh_neg,
#   verbose = TRUE
# )

#head(markers)

```


Markers are ranked based on:

* differential expression
* specificity
* co-expression with GNRH1


## Network visualization

The package also supports exploratory coexpression analysis.

Star Network

```{r}
#plot_star_network(df)
```

Similarity Network

```{r}
#plot_similarity_network(df)
```


## Diagnostic metadata

The pipeline stores rich metadata inside the Seurat object.

Available fields:

```{r}
#names(obj@meta.data)
```

Diagnostic layer:

```{r}
#head(obj@misc$gnrh_diag)
```

Threshold sweep:

```{r}
#head(obj@misc$gnrh_threshold_curve)
```

Stored parameters:

```{r}
#obj@misc$gnrh_params
```

# Recommended workflow

```{r}
# obj <- NormalizeData(obj)
# obj <- RunPCA(obj)
# obj <- RunUMAP(obj, dims = 1:30)
# obj <- cellgnrh(obj)
# obj <- gnrh_stage(obj)
# plot_gnrh_umap(obj)
# plot_gnrh_threshold(obj)
```


# Interpretation notes

High-confidence cells

gnrh_high cells represent the strictest classification
and are recommended for downstream biological analysis.

Low-confidence cells

gnrh_low cells may represent:

* transitional states
* partial lineage commitment
* ambient contamination
* low-depth sequencing

These should be interpreted cautiously.


Best Practices

Recommended settings for droplet-based scRNA-seq:

```{r}
# min_umi = 2
# mad_factor = 2
# score_q = 0.85
# min_counts = 500
```

For low-depth datasets:

```{r}
# min_counts = 200
# score_q = 0.75
```


## Cross-species transfer learning

The goal is to project GnRH lineage signatures learned in one species (human or mouse) onto another species using ortholog mapping and conserved module scoring.

A lightweight implementation can be fully compatible with Seurat.

```{r}
# query_obj = gnrh_transfer_learning(
#   ref_obj,
#   query_obj,
#   ref_species = "human",
#   query_species = "mouse",
#   ortholog_table = NULL,
#   assay = "RNA",
#   reduction = "cca",
#   dims = 1:30,
#   k.anchor = 5,
#   verbose = TRUE
# )
```


## Future extensions

Future versions of cellgnrh may incorporate machine learning and graph-based methods for improved rare-cell detection and developmental lineage modeling.

### Probabilistic classification

Current classification is based on adaptive thresholding of composite GnRH lineage scores. However, neuronal identity often exists along continuous developmental trajectories rather than discrete states.

Future implementations may therefore include probabilistic lineage inference models:

$P(\mathrm{GnRH}_i) = \frac{1}{1 + e^{-k(S_i - \mu)}}$

where:

- $(P(\mathrm{GnRH}i))$ is the probability that cell $(i)$ belongs to the GnRH lineage
- $(S_i)$ is the composite GnRH score
- $(\mu)$ is the score midpoint
- $(k)$ controls classification sharpness

Probabilistic modeling could improve:

- transitional state detection
- uncertainty estimation
- low-depth dataset analysis
- sparse transcriptomic classification

### Cross-species transfer learning

Because major developmental programs governing GnRH neuronal differentiation are evolutionarily conserved, future versions may support cross-species transfer learning.

This framework would project lineage signatures learned in one species (e.g. mouse) onto another species (e.g. human) using ortholog mapping and shared latent representations.

Potential applications include:

- comparative developmental neurobiology
- human–mouse lineage projection
- evolutionary conservation analysis
- annotation transfer across datasets

### Graph neural networks

Single-cell transcriptomic datasets are naturally represented as cell-cell neighborhood graphs. Future versions of cellgnrh may therefore incorporate graph neural network (GNN) architectures for neighborhood-aware lineage inference.

Graph-based models could leverage local transcriptomic topology to improve classification robustness under sparse expression conditions.

A graph convolutional framework may be formulated as:

$H_i^{(l+1)} = \sigma\left(\sum{j \in \mathcal{N}(i)}W^{(l)} H_j^{(l)}\right)$

where:

- $(H_i)$ represents the latent embedding of cell $(i)$
- $(\mathcal{N}(i))$ denotes neighboring cells
- $(W^{(l)})$ are learnable graph weights
- $(\sigma)$ is a nonlinear activation function

Such approaches may improve:

- developmental trajectory inference
- migratory lineage reconstruction
- neighborhood-aware classification
- rare-cell recovery
- continuous maturation modeling

Potential future implementations may include:

- graph convolutional networks (GCN)
- graph attention networks (GAT)
- variational graph autoencoders
- semi-supervised lineage propagation models



# Reproducibility

The workflow stores:

* thresholds
* module scores
* confidence scores
* diagnostic tables
* ROC metrics
* staging assignments

allowing full downstream reproducibility.




# Suggested methodological references

Single-Cell RNA-seq Analysis

* Stuart T, Butler A, et al. Comprehensive Integration of Single-Cell Data. Cell. 2019.
* Hafemeister C, Satija R. Normalization and variance stabilization of single-cell RNA-seq data using regularized negative binomial regression. Genome Biology. 2019.
* Lun ATL, McCarthy DJ, Marioni JC. A step-by-step workflow for low-level analysis of single-cell RNA-seq data. F1000Research. 2016.

Ambient RNA Correction

* Young MD, Behjati S. SoupX removes ambient RNA contamination from droplet-based single-cell RNA sequencing data. GigaScience. 2020.
* Yang S, et al. Decontamination of ambient RNA in single-cell RNA-seq with DecontX. Genome Biology. 2020.

Robust Statistics

* Rousseeuw PJ, Croux C. Alternatives to the Median Absolute Deviation. Journal of the American Statistical Association. 1993.

Developmental GnRH Biology

* Wray S. Development of gonadotropin-releasing hormone-1 neurons. Frontiers in Neuroendocrinology. 2001.
* Herbison AE. The Gonadotropin-Releasing Hormone Neuron and Its Role in the Control of Puberty. Endocrine Reviews. 2018.
* Cariboni A, Maggi R, Parnavelas JG. From nose to fertility: the long migratory journey of gonadotropin-releasing hormone neurons. Trends in Neurosciences. 2007.

# Citation

If you use cellgnrh in published work, please cite:

Mbouamboua Y. cellgnrh: Detection and Staging of GnRH-Expressing Cells from Single-Cell RNA-seq Data.

In addition, please cite the relevant biological and computational references listed above.


# Session information

```{r}
sessionInfo()
```

